evidence of adaptation-independent time-warp spike timing noise correlations in cortical networks

### 2020-08-18 manuscript.tex
Leaving library

% which may be multi-dimensional. 
Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.
Such changes in continuous cortical state can occur on long or short-timescales, and may be specific to neurons or neuron-types or general across the population.
% , and may occur across the population, or specifically for neurons or neuron-types (Fig. 0a). 
In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive} or changes in the level of anaesthesia. Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state. Analysis of neuronal responses over multiple trials may allow inference of a fluctuating shared short-timescale cortical state, with possible implications for neural coding.

@@

A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space. Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. The sample stimulus-response distribution is therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). For a single stimulus, trial-to-trial variability shared between multiple neurons (i.e. `noise correlations') corresponds with correlative structure in the stimulus response distribution\footnote{Noise correlations differ to `signal correlations', which refer to tuning similarity between neurons.}. With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses. If downstream neurons were also modulated by the same short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved coding capacity.

This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing noise correlations as a step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations (Fig. 1a). If the state fluctuations were low dimensional across a population, then it is more likely that decoding neurons within or outside the population would be modulated by the same fluctuations, and could therefore respond with improved coding capacity in a state-dependent manner.
Characterisation of noise correlations is complicated by changes to neural responses caused by longer-timescale processes. Such longer-timescale changes are therefore are controlled for in this paper, offering a prediction of encoding capacity under a state-dependent coding regime.

>>>>

% which may be multi-dimensional. 
Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.
Such changes in continuous cortical state can occur on long or short-timescales, and may be specific to neurons or neuron-types or general across the population (Fig. 1a).
% , and may occur across the population, or specifically for neurons or neuron-types (Fig. 0a). 
In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive} or changes in the level of anaesthesia (Fig. 1a). Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state (Fig. 1a). Analysis of neuronal responses over multiple trials may allow inference of a fluctuating shared short-timescale cortical state, with possible implications for neural coding (Fig. 1a).

@@

A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space (Fig. 1b). Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. The sample stimulus-response distribution is therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1b).

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1c). For a single stimulus, trial-to-trial variability shared between multiple neurons (`noise correlations'\footnote{Noise correlations differ to `signal correlations', which refer to tuning similarity between neurons.}) corresponds with correlative structure in the stimulus response distribution when the dimensions correspond to spike train metrics for different individual neurons (Fig. 1d). With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses. If downstream neurons were also modulated by the same short-timescale fluctuating state, then downstream `decoding' neurons could respond in a state-dependent manner.
When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1e). State-dependent decoding would therefore offer an improved coding capacity.

If the state fluctuations were low dimensional across a population, then it is more likely that decoding neurons within or outside the population would be modulated by the fluctuations modulating afferently connected encoding neurons, and could therefore respond with improved coding capacity in a state-dependent manner.
This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing noise correlations, towards a better understanding of stimulus representations and as an important step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations. 
Characterisation of noise correlations is complicated by longer-timescale changes to neural responses. Such longer-timescale changes are therefore are controlled for in this paper, offering a prediction of encoding capacity under a state-dependent coding regime.
